
# "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true


## Loading and preprocessing the data
### Loading the Data

```{r}
library(ggplot2)
library(plyr)
if(!file.exists('activity.csv')){
  unzip('activity.zip')
}
```
###Data Processing Step
```{r}
activity <- read.csv("activity.csv")
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime<- as.POSIXct(activity$date, format="%Y-%m-%d")
clean <- activity[!is.na(activity$steps),]
```


## What is mean total number of steps taken per day?

```{r}
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum, )
colnames(sumTable)<- c("Date", "Steps")


### making a histogram for number of steps taken on daily basis

hist(sumTable$Steps, breaks= 50, xlab="Steps", main = "Total Steps per Day", col = "orange")

### Steps' Mean
as.integer(mean(sumTable$Steps))
### Steps' Median
as.integer(median(sumTable$Steps))
```
### The Mean for number of steps taken per day is 10766
### The Median for number of steps taken per day is 10765




## What is the average daily activity pattern?

```{r}
clean <- activity[!is.na(activity$steps),]
intervalTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))

### Lineplot creation

p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")

### Intervalwise maximum steps

maxSteps <- max(intervalTable$Avg)

### Interval containing max avg number of steps

intervalTable[intervalTable$Avg==maxSteps,1]
```
### The 5-minute interval withe maximum number of steps is the 835th interval.





## Imputing missing values
```{r}

nrow(activity[is.na(activity$steps),])
```
### There are 2304 number of rows where steps has 'NA' value

### I would like to fill the NAs with NAs with the average 5-minute interval value based on the day of the week.

```{r}

### average number of steps per interval and weekday basis

avgTable <- ddply(clean, .(interval, day), summarize, Avg = mean(steps))
nadata<- activity[is.na(activity$steps),]
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
```

### CReating duplicate datasets with filled in missing values

```{r}
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")
mergeData <- rbind(clean, newdata2)
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(sumTable2)<- c("Date", "Steps")
as.integer(mean(sumTable2$Steps))
as.integer(median(sumTable2$Steps))

### Creating Histogram
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Red")

hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Yellow", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("Red", "yellow") )
```

### The overall Distribution shape is in unaltered state. However new mean is 10821 compared to 10766 (Old mean) and the new median is 11015 compared to 10765 (Old Median).



## Are there differences in activity patterns between weekdays and weekends?

```{r}
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
library(lattice) 
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```


### From the plot we can clearly see that there is a difference in activities of people between weekdays and weekends.
